Detection of eco-evolutionary dynamics in metacommunities using Joint Species

### 2024-11-11 manuscript/manuscript_v12.Rmd
Draft 12 with full Discussion

```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments. The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents. The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density. This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
knitr::include_graphics("../output/fig5.pdf")
```

@@

By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics. This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020] and does not require additional common garden experiments [@Garzon2019]. Therefore this allows for a straight forward implementation of HMSC on a variety of data structures as lone as a clear response variable (e.g. species density) and ecological (e.g. environmental conditions) as well as evolutionary (e.g. heritable trait change) are measured. An additional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015]. With the implementation this approach can be combined with the fixed effects to answer further questions, such as: How does trait change impact the species density under changing environmental conditions? And how does the impact of trait changes depend on specific environmental conditions?

@@

>>>>

```{r fig3, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments. The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents. The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density. This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
knitr::include_graphics("../output/fig5.pdf")
```

@@

*Main findings*
The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity. A more difficult, yet critical, element to incorporate is species interactions. Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; [@Poggiato2021]). Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments). We used an existing JSDM and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure. We also applied this statistical model (HMSC, hierarchical models of species distributions; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects. Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.

*Discussion regarding simple case studies*
At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution. For example, our simulations reveal instances where trait evolution may occur at the same time as large changes in population size, and this co-occurrence isn't always causal (e.g. when selection is weak but trait heritability is high, and species are introduced at low initial population size). In this instance, the regression coefficient from the model may be negative and strong. By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; \@ref(fig:fig1); Appendix S1), we can better interpret effect size estimates. However, these precautions are not unique to including trait evolution as a predictor of community composition. For example, previous research has shown that the effect size coefficients from linear and AR models using $N_{i,t-1}$ to predict $N_{i,t}$ or $N_{j,t}$ are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ([@Kloppers2013] [@Certain2018] [@Muhlbauer2020] [@Olivenca2021]). Our findings for using HMSC to estimate the impact of trait evolution ($|\Delta x_i|$) for species abundances are similar - (i) the relative importance of trait evolution estimated by a variation partition accurately reflects the relative strength of simulated species interactions vs. magnitude of trait change and (ii) the coefficients greatly increase the predictive ability of the statistical model for simulated data (Appendix S1). HMSC is therefore a useful tool for understanding the role that evolution can play in community dynamics.

*Discussion regarding metacom simulations*
This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure 2a \@ref(fig:fig1); Appendix S4). First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e. [@Gomulkiewicz1995]) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e. community rescue, [@Low2015] [@Bell2019]; Figure 2b \@ref(fig:fig2)). The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species. Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S4), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure. Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate $d$) and the overall speed of adaptive evolution (modeled as heritability $h^2$). Previous theoretical models have been shown that these two properties can have a strong influence on resulting community composition ([@Urban2008] [@Loeuille2008] [@Urban2009] [@DeMeester2016]), but our demonstration that the HMSC statistical model can accurately estimate $R^2$ values for the relative importance of trait evolution can allow hypotheses from theoretical models to be tested using empirical data and compared across different study systems.

*Discussion regarding empirical data*
By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics. This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020] and does not require additional common garden experiments [@Garzon2019]. Therefore this allows for a straight forward implementation of HMSC on a variety of data structures as lone as a clear response variable (e.g. species density) and ecological (e.g. environmental conditions) as well as evolutionary (e.g. heritable trait change) are measured. An additional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015]. With the implementation this approach can be combined with the fixed effects to answer further questions, such as: How does trait change impact the species density under changing environmental conditions? And how does the impact of trait changes depend on specific environmental conditions?

@@

*Takehome points and conclusions*
There has been a strong general interest in estimating effect sizes of evolution for ecological processes [@Hairston2005]; [@Ellner2011]; [@Pantel2015]; [@Barbour2022]), but in many empirical studies these estimates focus on one or a few species where the impacts of their evolution is strong enough to play a clear and detectable role in the corresponding ecological response variable of interest (e.g. population growth rate, community diversity, likelihood of food web collapse). There is no general knowledge for overall distributions of effect sizes for the impact of evolution for species composition in communities, nor are there studies that evaluate drivers of shifts in these distributions of effect sizes in complex communities (see [@Hermann2024] for estimates in a predator-prey system). We present here distributions of effect sizes of evolution for species abundances (i.e. Figure 1c,f,i \@ref(fig:fig1); Figure 2c \@ref(fig:fig2); Figure 3c \@ref(fig:fig3)) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.

An increasing number of studies collect and analyze high-resolution time series of populations and trait values for single [@Rudman2022] and multiple [@Thomas2018] [@Gibert2023] [@Han2023] [@Shen2023] species. With increasing access to ecological and evolutionary data comes an increase in the number of interacting processes that should be considered to identify causal mechanisms that structure this data [@Pantel2023]. HMSC, and joint species distribution models generally, can be expanded to further include dynamic trait change that occurs during the course of community ecological studies. Two critically important additions that will increase the application to field or experimental observation data will be (1) bidirectional causality and eco-evolutionary feedbacks and (2) the relative contribution of plasticity and genotypic evolution for trait change (i.e. [@Gibert2022]). It will also be necessary to improve knowledge of the distributions of eco-evolutionary effects and how these are influenced by causal processes that change the speed of evolution or the strength of species interactions. Although these interacting eco-evolutionary dynamic processes represent an increased degree of complexity for data analysis and interpretation, these mechanistic processes can be translated into simpler statistical models to estimate ecological and evolutionary drivers of community composition.
